First Order Alternation

### 2019-02-04 quantifiers/body.tex
ok

Traditional attempts to generalize classical Rabin-Scott automata to
infinite alphabets, such as timed automata \cite{AlurDill94} and
finite-memory automata \cite{KaminskiFrancez94} face the
\emph{complement closure} problem: there exist automata for which the
complement language cannot be recognized by an automaton in the same
class. This excludes the possibility of encoding a language inclusion
problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
denotes the complement of $\lang{B}$.

@@

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~lazy annotation procedure, originally developed for
checking safety of nondeterministic integer programs
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure unfolds an automaton $A$ trying to find an execution that
recognizes a word $w \in \lang{A}$. If a path that reaches a final
state does not correspond to a concrete run of the automaton, the
positions on the path are labeled with interpolants from the proof of
infeasibility, thus marking this path and all continuations as
infeasible for future searches.

Checking emptiness of first-order alternating automata by lazy
annotation with interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules make it hard, if not impossible, to
  decide if a path is infeasible. This is mainly because adding
  uninterpreted predicate symbols to decidable first-order theories,
  such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the infeasibility of a path are not
  \emph{local}, as they may refer to input values encountered in the
  past. However, the future executions are oblivious to \emph{when}
  these values have been seen in the past and depend only on the
  relation between the past and current values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

@@

We encode timed automata as first order alternating automata and uses
closure of the latter class under intersection and complement to
compute an alternating automaton whose emptiness proves the inclusion
$\lang{A} \subseteq \lang{B}$ between timed regular languages. The
idea is to represent each control state of $A$ and $B$ by a binary
predicate symbol with the same name. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}. The first argument of
$s_1(x,z)$ is used to track the value of the $x$ clock of $A$, whereas
$z$ is the current time value. The time stamp of the input event is
the input variable $t$, which is bigger or equal than the current time
$z$, since the lapse of time between two consequtive events is always
positive. The current value of the clock is $z-x$, thus the guard $x
\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
parameter of $s_2$, in the first rule. Note that the time increment
equals the increment of the clocks that are not reset.

@@

The work probably closest to ours concerns the model of predicate
automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Because thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.

@@

A \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
occurring in the right-hand side formula of a transition rule are
referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
is $\len{\A} \isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
  \arrow{a(X)}{} \psi \in \Delta}$.

The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
are the input data values that trigger the transition, whereas $q$ and
$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
in that state, respectively. Without loss of generality, we consider,
for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
most one such rule, as two or more rules can be joined using
disjunction.

The execution semantics of automata is given in close analogy with the
classical definition of propositional alternating automata, with
transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
propositional variable and $\phi$ a positive boolean combination of
propositional variables. For instance, $q_0 \arrow{a}{} q_1 \wedge q_2
\vee q_3$ means that the automaton can choose to transition in either
both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to saying
that a transition is a \emph{minimal model} of a right hand side
formula of a rule. In our example, both $\set{q_1 \leftarrow \top, q_2
  \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow
  \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models,
however $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
  \top}$, is a model but is not minimal. The original definition of
alternating finite-state automata \cite{ChandraKozenStockmeyer81}
works around this problem by considering boolean valuations (models)
instead of formulae. However, a finite description of a first order
alternating automaton must use formulae, because there can be
infinitely many satisfying interpretations that are incomparable
w.r.t. the pointwise subset ordering.

@@

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
cube $c$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
  is a transition rule of $\A$.
\end{compactenum}
\end{definition}
An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if all paths in $\T$ have the same length and the frontier of
each tree $T \in \T$ is labeled with final configurations.  If $\A$
has an accepting execution over $w$ starting with a cube $c \in
\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$. We consider the
following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following \emph{relative
  acceptance} problem first: given an automaton $\A$ and an input
sequence $\alpha \in \Sigma^*$, does there exists a word $w \in
\lang{\A}$ such that $\event{w} = \alpha$ ?  By solving this problem
first, we develop the machinery required to prove that first order
alternating automata are closed under all boolean operations and,
further, set up the ground for a semi-algorithm for the emptiness
problem.

\subsection{Path Formulae}

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events.  Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only accepting executions over words
that share a given input sequence. Otherwise stated,
$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
a word $w$ such that $\event{w} = \alpha$. We reduce thus the relative
acceptance problem to the satisfiability of an effectively computable
formula in the first order theory of data with uninterpreted predicate
symbols.

Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicate
symbols and variables, respectively. As a shorthand, we write
$\stamp{Q}{\leq n}$ (resp.  $\stamp{X}{\leq n}$) for the set
$\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i}
  \mid x \in X, i \in [n]}$). For a formula $\psi$ and $i \in \nat$,
we define $\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$
the formula in which all input variables and state predicates (and
only those symbols) are replaced by their time-stamped
counterparts. As a shorthand, we shall write $q(\vec{y})$ for
$q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.

@@

As an immediate consequence, one can decide whether $\A$ accepts some
word $w$ with a given input sequence $\event{w}=\alpha$, by checking
whether $\accform{\alpha}$ is satisfiable. However, unlike
non-alternating infinite-state models of computation, such as counter
automata (nondeterministic programs with integer variables), the
satisfiability query for an acceptance (path) formula falls outside of
known decidable theories, supported by standard SMT solvers. There are
basically two reasons for this, namely\begin{inparaenum}[(i)]
\item the presence of predicate symbols, and
\item the non-trivial alternation of quantifiers.
\end{inparaenum}
To understand this point, consider for example, the decidable theory
of Presburger arithmetic \cite{Presburger29}. Adding even only one
monadic predicate symbol to it yields undecidability in the presence
of non-trivial quantifier alternation \cite{Halpern91}. However the
quantifier-free fragment of Presburger arithmetic extended with
predicate symbols can be shown to be decidable, using a Nelson-Oppen
style congruence closure argument \cite{NelsonOppen80}.

@@

For the time being, let us formally define the elimination of
transition quantifiers and predicates, respectively. Consider a given
sequence of input events $\alpha = a_1 \ldots a_n$ and denote by
$\alpha_i$ the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
where $\alpha_0=\epsilon$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ifLongVersion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}\label{def:quantform}
  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
  \bigwedge_{\begin{array}{l}
      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
      \]  
      We write
    $\quantform{\alpha}$ for the prenex normal form of the formula: 
      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
          \scriptstyle{q \in Q \setminus F}
      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
\end{definition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\else
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
sequence of formulae defined by $\quantformn{\alpha_0} \isdef
\stamp{\iota}{0}$ and, for all $i \in [1,n]$, let
$\quantformn{\alpha_i}$ be the conjunction of
$\quantformn{\alpha_{i-1}}$ with all formulae
$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow
\stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, such that
$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ occurs in
$\quantformn{\alpha_{i-1}}$. Next, we write $\quantform{\alpha}$ for
the conjunction of $\quantformn{\alpha_n}$ with all
$\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot$, such that
$\stamp{q}{n}(t_1,\ldots,t_{\#(q)})$ occurs in
$\quantformn{\alpha_n}$, for some $q \in Q \setminus F$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fi
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Note that $\quantform{\alpha}$ contains no path quantifiers, as
required. On the other hand, the scope of the transition quantifiers
in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
transition rules, as shown by the following example.

@@

Finally, we define the acceptance of a word with a given input event
sequence by means of a quantifier free formula in which no predicate
atom occurs.

@@

Given a positive formula $\phi$, we define the \emph{dual} formula
$\dual{\phi}$ recursively as follows:
\[\begin{array}{rclcrclcrcl}
\dual{(\phi_1 \vee \phi_2)} & \isdef & \dual{\phi_1} \wedge \dual{\phi_2} && 
\dual{(\phi_1 \wedge \phi_2)} & \isdef & \dual{\phi_1} \vee \dual{\phi_2} &&
\dual{(t = s)} & \isdef & t \neq s \\
\dual{(\exists x ~.~ \phi_1)} & \isdef & \forall x ~.~ \dual{\phi_1} && 
\dual{(\forall x ~.~ \phi_1)} & \isdef & \exists x ~.~ \dual{\phi_1} && 
\dual{(t \neq s)} & \isdef & t = s \\
&&&& \dual{q(x_1,\ldots,x_{\#(q)})} & \isdef & q(x_1,\ldots,x_{\#(q)})
\end{array}\]
The following theorem shows closure of automata under all boolean
operations:

@@

\paragraph{\em Proof sketch.} (\ref{it1:thm:closure}) The construction of $\A_\cap$ 
and $\A_\cup$ resembles the standard intersection and union of
propositional alternating automata
\cite{Arxiv}. (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a

>>>>

Traditional attempts to generalize classical Rabin-Scott automata to
infinite alphabets, such as timed automata \cite{AlurDill94} and
finite-memory automata \cite{KaminskiFrancez94} face the
\emph{complement closure} problem: there exist automata for which the
complement language cannot be recognized by an automaton in the same
class. This makes it impossible to encode a language inclusion
problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
denotes the complement of $\lang{B}$.

@@

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~lazy annotation procedure, originally developed for
checking safety of nondeterministic integer programs
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure unfolds an automaton $A$ trying to find an execution that
recognizes a word from $\lang{A}$. If a path that reaches a final
state does not correspond to a concrete run of the automaton, the
positions on the path are labeled with interpolants from the proof of
infeasibility, thus marking this path and all continuations as
infeasible for future searches. Termination of lazy annotation
procedures is not guaranteed, but having a suitable coverage relation
between the nodes of the search tree may ensure convergence of some
real-life examples. Applying lazy annotation to first order
alternating automata faces two nontrivial
problems: \begin{compactenum}
\item Quantified transition rules make it hard, if not impossible, to
  decide if a path is infeasible. This is mainly because adding
  uninterpreted predicate symbols to decidable first-order theories,
  such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the infeasibility of a path are not
  \emph{local}, as they may refer to input values encountered in the
  past. However, the future executions are oblivious to \emph{when}
  these values have been seen in the past and depend only on the
  relation between the past and current values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

@@

Our approach is to encode timed automata as first order alternating
automata and use closure of the latter class under intersection and
complement to compute an alternating automaton, whose emptiness proves
that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}.  Each control state of $A$
and $B$ is represented by a binary predicate symbol with the same
name. The first argument of $s_1(x,z)$ tracks the last moment in time
when the $\mathsf{x}$ clock of $A$ was reset and $z$ is the current
time value. The time stamp of the input event $a$ is the input
variable $t$, which must be bigger or equal to $z$, as the lapse of
time between two consequtive events is always positive. The current
value of the $\mathsf{x}$ clock is $z-x$, thus the guard $\mathsf{x}
\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
parameter of $s_2$, in the first rule.

@@

The work probably closest to ours concerns the model of predicate
automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Because thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier-free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.

@@

A \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.  Without loss
of generality, we consider, for each predicate $q \in Q$ and each
input event $a \in \Sigma$, at most one such rule, as two or more
rules can be joined using disjunction. The quantifiers occurring in
the right-hand side formula of a transition rule are referred to as
\emph{transition quantifiers}. The \emph{size} of $\A$ is $\len{\A}
\isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
  \arrow{a(X)}{} \psi \in \Delta}$.

@@

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
cube $c$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
  is a transition rule of $\A$.
\end{compactenum}
\end{definition}
The execution semantics of automata is given in close analogy with the
classical definition of propositional alternating automata, with
transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
propositional variable and $\phi$ a positive boolean combination of
propositional variables. For instance, $q_0 \arrow{a}{} q_1 \wedge q_2
\vee q_3$ means that the automaton can choose to transition in either
both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to saying
that a transition is a \emph{minimal model} of a right hand side
formula of a rule. In our example, both $\set{q_1 \leftarrow \top, q_2
  \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow
  \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models,
however $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
  \top}$, is a model but is not minimal. The original definition of
alternating finite-state automata \cite{ChandraKozenStockmeyer81}
works around this problem by considering boolean valuations (models)
instead of formulae. In contrast, a finite description of a first
order alternating automaton must be symbolic (in terms of formulae)
because here, a given configuration can have infinitely many
successors.

An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if all paths in $\T$ have the same length and the frontier of
each tree $T \in \T$ is labeled with final configurations.  If $\A$
has an accepting execution over $w$ starting with a cube $c \in
\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$. We consider the
following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following problem next: given an
automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
solving this problem first, we develop the machinery required to prove
that first order alternating automata are closed under all boolean
operations and, further, set up the ground for a semi-algorithm for
the emptiness problem.

\subsection{Path Formulae}
\label{sec:path-formulae}

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events.  Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only accepting executions over words
that share a given input sequence, i.e.\ $\accform{\alpha}$ is
satisfiable if and only if the automaton accepts a word $w$ such that
$\event{w} = \alpha$.

Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicate
symbols and variables, respectively. We also define $\stamp{Q}{\leq n}
\isdef \set{\stamp{q}{i} \mid q \in Q, i \in [n]}$ and $\stamp{X}{\leq
  n} \isdef \set{\stamp{x}{i} \mid x \in X, i \in [n]}$. For a formula
$\psi$ and $i \in \nat$, we define $\stamp{\psi}{i} \isdef
\psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the formula in which all input
variables and state predicates (and only those symbols) are replaced
by their time-stamped counterparts. Moreover, we write $q(\vec{y})$
for $q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.

@@

As an immediate consequence, one can decide whether $\A$ accepts some
word $w$ with a given input sequence $\event{w}=\alpha$, by checking
whether $\accform{\alpha}$ is satisfiable. However, unlike
non-alternating infinite-state models of computation, such as counter
automata (nondeterministic programs with integer variables), the
satisfiability query for an acceptance (path) formula falls outside of
known decidable theories, supported by standard SMT solvers. There are
basically two reasons for this, namely\ \begin{inparaenum}[(i)]
\item the presence of predicate symbols, and
\item the non-trivial alternation of quantifiers.
\end{inparaenum}
To understand this point, consider for example, the decidable theory
of Presburger arithmetic \cite{Presburger29}. Adding even only one
monadic predicate symbol to it yields undecidability in the presence
of non-trivial quantifier alternation \cite{Halpern91}. However the
quantifier-free fragment of Presburger arithmetic extended with
predicate symbols can be shown to be decidable, using a Nelson-Oppen
style congruence closure argument \cite{NelsonOppen80}.

@@

For the time being, let us formally define the elimination of
transition quantifiers and predicates, respectively. Let $\alpha = a_1
\ldots a_n$ be a given sequence of input events and let
$\alpha_i$ be the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
where $\alpha_0=\epsilon$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ifLongVersion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}\label{def:quantform}
  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
  \bigwedge_{\begin{array}{l}
      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
      \]  
      We write
    $\quantform{\alpha}$ for the prenex normal form of the formula: 
      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
          \scriptstyle{q \in Q \setminus F}
      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
\end{definition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\else
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider the sequence of formulae $\quantformn{\alpha_0}, \ldots,
\quantformn{\alpha_n}$ defined as $\quantformn{\alpha_0} \isdef
\stamp{\iota}{0}$ and, for all $i \in [1,n]$, let
$\quantformn{\alpha_i}$ be the conjunction of
$\quantformn{\alpha_{i-1}}$ with all formulae
$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow
\stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, such that
$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ occurs in
$\quantformn{\alpha_{i-1}}$. Next, we write $\quantform{\alpha}$ for
the conjunction of $\quantformn{\alpha_n}$ with all
$\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot$, such that
$\stamp{q}{n}(t_1,\ldots,t_{\#(q)})$ occurs in
$\quantformn{\alpha_n}$, for some $q \in Q \setminus F$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fi
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Note that $\quantform{\alpha}$ contains no path quantifiers, as
required. On the other hand, the scope of the transition quantifiers
in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
transition rules, as shown by the following example.

@@

Finally, we define the acceptance of a word with a given input event
sequence by means of a quantifier-free formula in which no predicate
atom occurs.

@@

Given a positive formula $\phi$, we define the \emph{dual} formula
$\dual{\phi}$ recursively as follows:
\[\begin{array}{rclcrclcrcl}
\dual{(\phi_1 \vee \phi_2)} & \isdef & \dual{\phi_1} \wedge \dual{\phi_2} && 
\dual{(\phi_1 \wedge \phi_2)} & \isdef & \dual{\phi_1} \vee \dual{\phi_2} &&
\dual{(t = s)} & \isdef & t \neq s \\
\dual{(\exists x ~.~ \phi_1)} & \isdef & \forall x ~.~ \dual{\phi_1} && 
\dual{(\forall x ~.~ \phi_1)} & \isdef & \exists x ~.~ \dual{\phi_1} && 
\dual{(t \neq s)} & \isdef & t = s \\
&&&& \dual{q(x_1,\ldots,x_{\#(q)})} & \isdef & q(x_1,\ldots,x_{\#(q)})
\end{array}\]
The following theorem shows closure of automata under all boolean
operations. Note that it is sufficient to show closure under
intersection and negation because $\lang{\A_1} \cup \lang{\A_2} =
\Sigma[X]^* (\clang{\A_1} \cap \clang{\A_2})$, for any two automata
$\A_1$ and $\A_2$ with input event alphabet $\Sigma$ and input
variables $X$.

@@

\paragraph{\em Proof sketch.} (\ref{it1:thm:closure})
The proof for $\A_\cap$ resembles the standard proof for the
intersection of propositional alternating automata
\cite{IosifXu18Arxiv} (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$
be a word. We denote by $\accformA{\A_1}{\event{w}}$ and
